---
title: "Years as coloured bars: better alternatives using ggplot2"
author: "Neil Saunders"
date: 'compiled `r Sys.time()`'
output: 
  github_document: 
    toc: true
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE)
library(dplyr)
library(ggplot2)

theme_set(theme_light())

set.seed(1001)
example_data2 <- data.frame(
  category = rep(paste0("Category", 1:6), each = 6),
  year = rep(2000:2005, by = 6),
  n = sample(1:30, 36, replace = TRUE)
)
```

# Introduction
I often see charts with "years as coloured bars". What's better than years as coloured bars?

# Dataset
We generate some fake categories over time.

## Bars coloured by year
This is what we don't want: years as coloured bars within categories.

```{r bars-year}
example_data2 %>%
  ggplot(aes(category, n)) + 
  geom_col(aes(fill = factor(year)), position = "dodge") + 
  scale_fill_brewer(palette = "RdYlBu", name = "year")
```

Bars in one colour with no legend? No (not shown).

```{r bars-year-mono, eval=FALSE}
example_data2 %>%
  ggplot(aes(category, n)) + 
  geom_col(aes(fill = factor(year)), position = "dodge") + 
  scale_fill_manual(values = rep("#91BFDB", 6)) +
  guides(fill = FALSE)
```

## Bars coloured by category
Switching categories for years makes it difficult to follow individual categories over time.

```{r bars-category}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_col(aes(fill = factor(category)), position = "dodge") + 
  scale_fill_brewer(palette = "RdYlBu", name = "category")
```

## Bars coloured by category, stacked
Stacking makes it difficult to follow categories when a value changes abruptly.

```{r bars-category-stacked}
example_data2 %>%
  ggplot(aes(year, n)) + 
  geom_col(aes(fill = factor(category)), position = "stack") + 
  scale_fill_brewer(palette = "RdYlBu", name = "category")
```

## Bars coloured by category, stacked and filled
Stacking and filling creates confusion when values change for any category other than the bottom row.

```{r bars-category-filled}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_col(aes(fill = factor(category)), position = "fill") + 
  scale_fill_brewer(palette = "RdYlBu", name = "category")
```

## Area coloured by category, stacked and filled
Area plots help somewhat, but make it difficult to follow changes in absolute value per category.

```{r area-category-filled}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_area(aes(fill = factor(category)), position = "fill") + 
  scale_fill_brewer(palette = "RdYlBu", name = "category")
```

## Lines coloured by category
Overlapping lines are messy.

```{r lines-category}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_line(aes(color = factor(category))) + 
  scale_color_brewer(palette = "RdYlBu", name = "category")
```

## Facets
### Facet grid by category, bars
Facet by category is better but bars are disjointed.

```{r bars-category-facet}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_col(fill = "#91BFDB") + 
  facet_grid(category ~ .)
```

### Facet grid by category, lines
Lines are better but there is still some visual distraction from the other categories.

```{r lines-category-facet}
example_data2 %>% 
  ggplot(aes(year, n)) + 
  geom_line() + 
  facet_grid(category ~ .)
```

### Facet wrap by year, bars
This clearly shows categories per year, but we want changes in each category over time.
```{r bars-year-facet}
example_data2 %>%
  ggplot(aes(category, n)) +
  geom_col(fill = "#91BFDB") + 
  facet_wrap(~year) + 
  theme(axis.text.x = element_text(angle = 90))
```

### Facet wrap by category, lines
Success! "Years within categories" was correct after all - we just needed to facet and use lines.

```{r lines-year-facet}
example_data2 %>%
  ggplot(aes(year, n)) + 
  geom_line() + 
  facet_wrap(~category) +
  theme(panel.spacing.x = unit(1, "lines"))
```
