---
title: "Exploring Anscombe's quartet using R + ggplot2"
author: "Neil Saunders"
date: "`r Sys.time()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

library(tidyverse)
library(gridExtra)

theme_set(theme_bw())

data("anscombe")
```

# Introduction
[Anscombe's quartet](https://en.wikipedia.org/wiki/Anscombe's_quartet) is a set of four small datasets, constructed to show the importance of visualising data and the dangers of reliance on simple summary statistics.

# The data
As it is a "classic" dataset, the quartet is included in the R `datasets` package. You can type `?anscombe` for the details, which include example code for analysis and plotting.

First, load required libraries and data.

The `summary` function illustrates some of the identical and near-identical properties of the data. It's also easy to confirm the values for variance, correlation coefficient and linear regression given in the Wikipedia article.

```{r summary}
summary(anscombe)

# correlation
sapply(1:4, function(x) cor(anscombe[, x], anscombe[, x+4]))

# variance
sapply(5:8, function(x) var(anscombe[, x]))

# linear regression (first pair only)
lm(y1 ~ x1, data = anscombe)
```

# Plotting the data
Using `ggplot2` and `gridExtra` to create plots similar to those in the Wikipedia article.

```{r plots, fig.width=9, fig.height=7}
p1 <- anscombe %>%
  ggplot(aes(x1, y1)) + 
    geom_point(color = "darkorange", size = 3) + 
    scale_x_continuous(breaks = seq(0, 20, 2)) + 
    scale_y_continuous(breaks = seq(0, 12, 2)) + 
    geom_abline(intercept = 3, slope = 0.5, color = "cornflowerblue") + 
    expand_limits(x = 0, y = 0) + 
    labs(title = "dataset 1")

p2 <- anscombe %>%
  ggplot(aes(x2, y2)) + 
    geom_point(color = "darkorange", size = 3) + 
    scale_x_continuous(breaks = seq(0, 20, 2)) + 
    scale_y_continuous(breaks = seq(0, 12, 2)) + 
    geom_abline(intercept = 3, slope = 0.5, color = "cornflowerblue") + 
    expand_limits(x = 0, y = 0) + 
    labs(title = "dataset 2")

p3 <- anscombe %>%
  ggplot(aes(x3, y3)) + 
    geom_point(color = "darkorange", size = 3) + 
    scale_x_continuous(breaks = seq(0, 20, 2)) + 
    scale_y_continuous(breaks = seq(0, 12, 2)) + 
    geom_abline(intercept = 3, slope = 0.5, color = "cornflowerblue") + 
    expand_limits(x = 0, y = 0) + 
    labs(title = "dataset 3")

p4 <- anscombe %>%
  ggplot(aes(x4, y4)) + 
    geom_point(color = "darkorange", size = 3) + 
    scale_x_continuous(breaks = seq(0, 20, 2)) + 
    scale_y_continuous(breaks = seq(0, 12, 2)) + 
    geom_abline(intercept = 3, slope = 0.5, color = "cornflowerblue") + 
    expand_limits(x = 0, y = 0) + 
    labs(title = "dataset 4")

grid.arrange(p1, p2, p3, p4, top = "Anscombe's Quartet")
```
